Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-10 paper.tex
More discussion

\section{Discussion}
VCF is a central element of modern genomics, facilitating
the exchange of data among interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the current generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
and likely environmental cost~\cite{grealey2022carbon}.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a solution to all problems. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of various fields are determined by the \emph{maximum}
dimension among all variants. Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider problems of ongoing updates
(i.e, as new samples and their variants are added).
Thirdly, Zarr is not suitable for handling unstructured
string data, and for example likely not work well as a 
basis for storing annotation data [CITES].

% Perhaps the largest deviation in the approach that we advocate here 
% from standard practice is to decouple our concept of a dataset from
% a file. Each VCF is broken up into a set of Zarr arrays, each 
% corresponding to a large number of chunks stored in a directory
% hierarchy (Zarr also supports other forms of chunk storage 
% including single files and key-value stores however; see Methods).
% While there are some disadvantages to large numbers of small files,
% we argue that the benefits greatly outweigh them, particularly 
% in the large, centrally controlled dataset setting. 
% Firstly, this approach is inherently parallelisable, and chunks across the 
% dataset written and read in parallel without needing complex 
% synchronisation and prior layout planning. 
% Such parallel write access is very difficult 
% in the single-file setting, and one of the key weaknesses 
% of the HDF5 format~\citep{folk2011overview} [CITATION with 
% parallel troubles from HDF5?]
% Secondly, mapping chunks in arrays to individually addressed
% keys maps well to large-scale cloud object storage. Indeed,
% Zarr is specifically designed for such usage, and 
% several petabyte scale datasets have been successfully deployed
% and used [citations?].
% We have focused here on the classical single-server and 
% HPC cluster settings, but a major advantage of Zarr is 
% its cloud readiness.

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]

@@

The second key piece of infrastructure required at this large 
scale concerns distributed computation. Even with more efficient
columnar data representations, computations that required examing 
large volumes of per-call data must be distributed to complete 
in a reasonable amount of time. [Dask, Xarray, sgkit?]
Currently users are responsible for orchestrating such calculations
on clusters, and automation would be a significant boon
for reasearcher productivity. [Hail has demonstrated the value of this?]

% We suggest that the future of VCF is to explicitly define the underlying
% data model, and to decouple it from the current row-wise encodings 
% of the VCF text and BCF binary file formats. This will enable data 
% to flow more freely, and provide scope for much-needed innovation.

% \item A major strong point of Zarr is its simplicity. All it's doing
% is compressing n-dimensional chunks of numerical data and storing
% them using fixed address keys. This key-value approach means that storage
% is very flexible and implementation is straightforward. There
% are several implementations of the Zarr protocol, and while some of these
% are rudimentary, they could be improved or replaced with relatively
% little effort. Zarr has been successfully used to store petabytes
% of data across different sciences --- genetic variation data is largely
% arrays of numerical data, and not really that different ultimately.
% Zarr is cloud native.
% Zarr is an open specification, built on standard components (contrast
% with Hail).

@@

% Workflow engines such as
% Snakemake~\cite{koster2012snakemake,molder2021sustainable}
% and [TODO mention WDL, Cromwell, etc?] make such analyses
% much easier, but there is still a great deal of room for
% mistakes.
% Per-chromosome VCFs are not practical for biobank scale
% projects, however, and variant call data tends to be split
% into many (still large) files.
% For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
% are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
% hundreds of files per chromosome.
% % This is horrible writing, but the point is important.
% While these files provide a natural
% unit for distributing work, the details of how they
% are split differ and essentially there is
% no interoperability among large-scale datasets because the VCFs are
% so cumbersome that a separate orchestration layer is required to
% access the data in parallel.

>>>>

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data among interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the current generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
and likely environmental cost~\cite{grealey2022carbon}.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a solution to all problems. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of various fields are determined by the \emph{maximum}
dimension among all variants. Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider problems of ongoing updates
(i.e, as new samples and their variants are added).
Thirdly, Zarr is not suitable for handling unstructured
string data, and for example likely not work well as a 
basis for storing annotation data [CITES].

Nonetheless, there are numerous datasets that exist today
what would benefit from cloud-native storage, 
inherent parallelism and granular storage.
[Mention benefits of fine-grained access control to 
different parts of the dataset using native storage access
control mechanism?]

@@

The second key piece of infrastructure required at this large 
scale concerns distributed computation. Even with more efficient
columnar data representations, computations that required examing 
large volumes of per-call data must be distributed to complete 
in a reasonable amount of time. [Dask, Xarray, sgkit?]

@@
